---
title: "Data Processing"
author: "Nasim Bondar Sahebi"
subtitle: "[EcoTyper_Project, Data processing](https://github.com/sogolsahebi/EcoTyper_Project)"
date: "2024-11-01"
output:
  html_document: default
params:
  show_code: true  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## load libraries

```{r libraries}
library(SummarizedExperiment)
library(MultiAssayExperiment)
library(dplyr)
library(knitr)
library(kableExtra)
library(tools)
library(ggplot2)
library(networkD3)
```

## Data Loading and Preparation

This section includes immunotherapy datasets curated in the BHK Lab, with sources available on [GitHub](https://github.com/bhklab/ClinSets/tree/main/data). The datasets are organized as follows:

- **All Data**: Stored in the `data_all` folder, containing `.rda` files for each cohort (e.g., `ICB_Braun__Kidney__PD-(L)1.rda`).
- **Data Stratified by Sex**: Stored in the `data_sex` folder, with datasets split by gender. Files for female and male data are indicated with `_F` and `_M` in the filename (e.g., `ICB_Gide__Melanoma__PD-(L)1__F.rda` for female data and `ICB_Gide__Melanoma__PD-(L)1__M.rda` for male data).

This section will list and display all `.rda` files from the `data_all` and `data_sex` directories, covering a total of 56  study cohorts.

```{r  all data}

# Set base directory
Base_dir <- "~/BHK lab/Ecotyper_Project/"

# List .rda files in both directories
data_files <- list.files(path = file.path(Base_dir, "data_all"), pattern = "\\.rda$", full.names = TRUE)
data_sex_files <- list.files(path = file.path(Base_dir, "data_sex"), pattern = "\\.rda$", full.names = TRUE)

# Create a table of all studies
data_files_df <- data.frame(Files = basename(data_files))
kable(data_files_df, caption = "All studies", col.names = "Cohorts") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")


# Initialize summary table
summary_table <- data.frame(Dataset = character(),
                            Cancer_Type = character(),
                            Clinical_Endpoints = character(),
                            Patients = integer(),
                            Treatments = character(),
                            stringsAsFactors = FALSE)

# Loop to extract information from each dataset
for (file in data_files) {
  dat_icb <- get(load(file))
  dataset_name <- sub("ICB_", "", basename(file))
  dataset_name <- sub("\\.rda$", "", dataset_name)
  clin <- data.frame(colData(dat_icb))
  
  # Extract details
  cancer_type <- paste(unique(clin$cancer_type), collapse = ", ")
  clinical_endpoints <- paste(ifelse(unique(clin$survival_type) == "None", "Response (R vs NR)", 
                            unique(clin$survival_type)), collapse = ", ")
  
  patients <- dim(clin)[1]
  
  treatments <- paste(unique(clin$treatment), collapse = ", ")

  
  # Append to summary table
  summary_table <- rbind(summary_table, data.frame(Dataset = dataset_name,
                                                   Cancer_Type = cancer_type,
                                                   Clinical_Endpoints = clinical_endpoints,
                                                   Patients = patients,
                                                   Treatments = treatments,
                                                   stringsAsFactors = FALSE))

}

# Display summary table
kable(summary_table, caption = "Summary of Immunotherapy Datasets") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")


write.csv(summary_table, file = file.path(Base_dir, "outputs/Dataprocessing_outputs", "Summary_of_Each_Dataset_IO_Datasets.csv"), row.names = TRUE)

```

## Combine all Datasets

This section loads and combines clinical datasets from the `data` and `data_sex` directories. Key steps include:

- Listing `.rda` files from both directories.
- Extracting and merging clinical data based on common columns.
- Creating unique study and patient identifiers.
- Displaying combined clinical data in tables.

The final datasets are saved as `.rds` files for future use.

```{r Combine clinicaldata 1 }

# List .rda files in both directories
data_files <- list.files(path = file.path(Base_dir, "data_all"), pattern = "\\.rda$", full.names = TRUE)
data_sex_files <- list.files(path = file.path(Base_dir, "data_sex"), pattern = "\\.rda$", full.names = TRUE)

# Initialize lists for storing column names and clin dataframes for data_files
column_list <- list()
clin_list <- list()

for (data_path in data_files) {
    obj <- load(data_path)  
    se <- get(obj)

    clin <- as.data.frame(colData(se))  
    
    # Set row names with base name and index
    base_name <- tools::file_path_sans_ext(basename(data_path))
    rownames(clin) <- paste0(base_name, "_", seq_len(nrow(clin)))
    
    column_list[[data_path]] <- colnames(clin)  # Store column names
    clin_list[[data_path]] <- clin  # Store clinical data
}

# Identify common columns across all datasets
common_columns <- Reduce(intersect, column_list)

# Bind all clin dataframes using the common columns
combined_clin <- do.call(rbind, lapply(clin_list, function(clin) {
    clin <- data.frame(clin)
    clin[, common_columns, drop = FALSE]  # Subset to common columns
}))

combined_clin <- do.call(rbind, lapply(clin_list, function(clin) data.frame(clin)[, common_columns, drop = FALSE]))

# Create study and patient identifiers
combined_clin$study <- gsub("\\..*","", basename(rownames(combined_clin)))
combined_clin$study_patient <- paste0(combined_clin$study, "_", combined_clin$patientid)
rownames(combined_clin) <- combined_clin$study_patient

# Repeat for data_sex_files
data_sex_files_df <- data.frame(Files = basename(data_sex_files)) # List all data_sex .rda files

# Initialize lists for storing column names and clin dataframes for data_sex
column_sex_list <- list()
clin_sex_list <- list()

column_sex_list <- list()  # Initialize list for column names
clin_sex_list <- list()    # Initialize list for clinical data

for (data_sex_path in data_sex_files) {
    obj <- load(data_sex_path )  
    clin_sex <- data.frame(colData(get(obj)))
    
    # Set row names with base name and index
    base_name <- tools::file_path_sans_ext(basename(data_sex_path))
    rownames(clin_sex) <- paste0(base_name, "_", seq_len(nrow(clin_sex)))
    
    column_sex_list[[data_sex_path ]] <- colnames(clin_sex)  # Store column names
    clin_sex_list[[data_sex_path]] <- clin_sex  # Store clinical data
}

# Identify common columns across all data_sex datasets
common_sex_columns <- Reduce(intersect, column_sex_list)

# Fully flatten clin_sex_list and bind using common columns
combined_clin_sex <- do.call(rbind, lapply(clin_sex_list, function(clin_sex) {
    #clin_sex <- as.data.frame(clin_sex)
    clin_sex[, common_sex_columns, drop = FALSE]
}))

# Save as csv files
# write.csv(combined_clin, file = file.path(Base_dir, "outputs/Dataprocessing_outputs", "combined_clin_before_exclusion.csv"), row.names = TRUE)

# write.csv(combined_clin_sex, file = file.path(Base_dir, "outputs/Dataprocessing_outputs", "combined_clin_sex_before_exclusion.csv"), row.names = TRUE)

# Create study and patient identifiers for data_sex
combined_clin_sex <- combined_clin_sex[!is.na(combined_clin_sex$patientid), ]
combined_clin_sex$study <- gsub("\\..*","", basename(rownames(combined_clin_sex)))
combined_clin_sex$study_patient <- paste0(combined_clin_sex$study, "_", combined_clin_sex$patientid)
rownames(combined_clin_sex) <- combined_clin_sex$study_patient

# Display the combined_clin table with scrolling enabled
kable(head(combined_clin), caption = "Summary of Combined Clinical Data (Data Directory)", 
      col.names = colnames(combined_clin), format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")

```

## Summary of All Data and Data with Sex

This section generates a summary table of clinical characteristics from the combined clinical datasets. The table includes the number of studies, patients, age (with median and IQR), missing values (NA %), sex distribution, cancer types, treatments, response rates, and survival outcomes (PFS and OS). The function `generate_all_data()` processes the data and computes these metrics for both the full dataset and the dataset with sex-specific information.

### Exclusion Criteria:

We apply the following exclusions:

- Remove rows with the following treatments: `"IO+chemo"`, `"IO+targeted"`, `"chemo"`, `"chemo+targeted"`, and `"targeted"`.In other words only keep `"PD-1/PD-L1"`, `"IO+combo"`, or `"CTLA4"`.
- Remove dataset with fewer than 25 patients.
- Exclude the cancer type `"Lymph_node"`.
- Only include patients from the studies `"Gide"`, `"Riaz"`, and `"Abacus"`, and only include `"PRE"` samples.
- Only keep patients with RNA-seq data (exclude microarray  and NA patients)
- only keeping PRE samples 


```{r Table 1 }
# Function to apply exclusions and generate summary data

# treatments: "PD-1/PD-L1", "CTLA4", "IO+combo", "chemo", "IO+targeted", "targeted", "IO+chemo","chemo+targeted"
combined_clin_before <- combined_clin
combined_clin_sex_before <- combined_clin_sex 

# Remove dataset with fewer than 25 patients
datasets_25_or_more <- paste0("ICB_", summary_table$Dataset[summary_table$Patients >= 25])
combined_clin <- combined_clin[combined_clin$study %in% datasets_25_or_more, ] # 16 Dataset are excluded

# Remove rows with treatments "IO+chemo", "IO+targeted", "chemo", "chemo+targeted", and "targeted"
combined_clin <- combined_clin[!combined_clin$treatment %in% c("IO+chemo", "IO+targeted", "chemo", "chemo+targeted", "targeted"), ] # 11 dataset excluded

combined_clin_sex <- combined_clin_sex[!combined_clin_sex$treatment %in% c("IO+chemo", "IO+targeted", "chemo", "chemo+targeted", "targeted"), ]

# # Remove cancer types with less than 30 patients
# cancer_types_less_30 <- names(table(combined_clin$cancer_type)[table(combined_clin$cancer_type) < 30])
# combined_clin <- combined_clin[!combined_clin$cancer_type %in% cancer_types_less_30, ]
# combined_clin_sex <- combined_clin_sex[!combined_clin_sex$cancer_type %in% cancer_types_less_30, ]

# Remove "Lymph_node" cancer type
combined_clin <- combined_clin[combined_clin$cancer_type != "Lymph_node", ] # one dataset exluded
combined_clin_sex <- combined_clin_sex[combined_clin_sex$cancer_type != "Lymph_node", ]

# Subset only patients with RNA-seq data (exclude microarray and NA patients)
# RNA-seq data types: "rnaseq", "tpm", "fpkm"
# Remove rows with "microarray" in the rna column and update rna to "rnaseq" for the remaining rows
filtered_combined_clin <- combined_clin[combined_clin$rna != "microarray", ]
filtered_combined_clin$rna <- "rnaseq"

# Remove rows with "microarray" in the rna column and update rna to "rnaseq" for the remaining rows
filtered_combined_clin_sex <- combined_clin_sex[combined_clin_sex$rna != "microarray", ]
filtered_combined_clin_sex$rna <- "rnaseq"

# Save as csv files
#write.csv(filtered_combined_clin, file = file.path(Base_dir, "outputs/Dataprocessing_outputs", "combined_clin_after_exclusion.csv"), row.names = TRUE)

#write.csv(combined_clin_sex, file = file.path(Base_dir, "outputs/Dataprocessing_outputs", "combined_clin_sex_after_exclusion.csv"), row.names = TRUE)

# Function to extract the base of the study name, ignoring the "_F" and "_M" ( for counting data sex with same study name)
get_study_base <- function(study_name) {
  return(sub("_[FM]$", "", study_name))
}

# Generate data function with only the common treatments and cancer types
generate_all_data <- function(data) {
  
  # Extract unique cancer types and treatments from the provided data
  cancer_types <- unique(data$cancer_type)
  treatments <- unique(data$treatment)
  
  # Dynamically create the Characteristic vector
  Characteristic <-c(
    "N째 studies", "N째 patients", "Age year", "Median (IQR)", "NA (%)", "Sex",
    "Female (%)", "Male (%)", "NA (%)", "Cancer Type",
    paste0(cancer_types, " (%)"),  # Dynamically add cancer types
    "Treatment",
    paste0(treatments, " (%)"),    # Dynamically add treatments
    "Response", "No-response (%)", "Response (%)", "NA (%)",
    "PFS", "No-progression (%)", "Progressed (%)", "NA (%)",
    "Median PFS", "IQR PFS", "OS", "Alive (%)", "Dead (%)", "NA (%)",
    "Median OS", "IQR OS")
  
  data$study_base <- sapply(data$study, get_study_base)
   
  All_DATA <- c(
    length(unique(data$study_base)),    # N째 studies          
    length(data$patientid),                    # N째 patients          
    "",                                       # Age year placeholder
    
    # 1. Age year placeholder
    paste(median(as.numeric(data$age), na.rm = TRUE), " (", min(as.numeric(data$age), na.rm = TRUE), "-", max(as.numeric(data$age), na.rm = TRUE), ")", sep = ""),
    paste(sum(is.na(data$age)), " (", round(sum(is.na(data$age)) / nrow(data) * 100, 1), "%)", sep = ""),
        
    "",                                     
    
    # 2. Sex placeholder
    paste(sum(data$sex == "F", na.rm = TRUE) , " (",round(sum(data$sex == "F", na.rm = TRUE) / nrow(data) * 100, 1), "%)"),  # Female percentage
    paste( sum(data$sex == "M", na.rm = TRUE), " (",round(sum(data$sex == "M", na.rm = TRUE) / nrow(data) * 100, 1), "%)"),  # Male percentage
    paste(sum(is.na(data$sex)) , " (",round(sum(is.na(data$sex)) / nrow(data) * 100, 1), "%)"),  # NA percentage for sex
    "",                                       # Cancer Type placeholder
    
    # 3. Calculate common cancer type percentages
    sapply(cancer_types, function(ct) {
      paste(sum(data$cancer_type == ct), " (", round(sum(data$cancer_type == ct) / nrow(data) * 100, 1), "%)", sep = "")
    }), 
    "",                                       
    
    # 4. Treatment placeholder for common treatments
    sapply(treatments, function(tr) {
      paste(sum(data$treatment == tr), " (", round(sum(data$treatment == tr) / nrow(data) * 100, 1), "%)", sep = "")
    }), 
    "",                                      
    
    # 5. Response placeholder
    paste(sum(data$response == "NR", na.rm = TRUE), " (", round(sum(data$response == "NR", na.rm = TRUE) / nrow(data) * 100, 1), "%)"), 
    paste(sum(data$response == "R", na.rm = TRUE), " (",round(sum(data$response == "R", na.rm = TRUE) / nrow(data) * 100, 1), "%)"),
    paste(sum(is.na(data$response)), "(", round(sum(is.na(data$response)) / nrow(data) * 100, 1), "%)"),  
    "",
    
    # PFS
    paste(sum(data$event_occurred_pfs == 0, na.rm = TRUE), " (",
          round(sum(data$event_occurred_pfs == 0, na.rm = TRUE) / nrow(data) * 100, 1), "%)"), 
    paste(sum(data$event_occurred_pfs == 1, na.rm = TRUE), "(",
          round(sum(data$event_occurred_pfs == 1, na.rm = TRUE) / nrow(data) * 100, 1), "%)"),
    paste(sum(is.na(data$event_occurred_pfs)), " (",
          round(sum(is.na(data$event_occurred_pfs)) / nrow(data) * 100, 1), "%)"), 
    paste(round(median(data$survival_time_pfs, na.rm = TRUE), 1)),  # Median PFS
    round(IQR(data$survival_time_pfs, na.rm = TRUE), 1),              # IQR PFS

    "",                                       
    
    # 6. OS placeholder
    paste(sum(data$event_occurred_os == 0, na.rm = TRUE), " (",
          round(sum(data$event_occurred_os == 0, na.rm = TRUE) / nrow(data) * 100, 1), "%)"),  # Alive percentage for OS
    paste(sum(data$event_occurred_os == 1, na.rm = TRUE), " (",
          round(sum(data$event_occurred_os == 1, na.rm = TRUE) / nrow(data) * 100, 1), "%)"),  # Dead percentage for OS
    paste(sum(is.na(data$event_occurred_os)), " (",
          round(sum(is.na(data$event_occurred_os)) / nrow(data) * 100, 1), "%)"),  # NA percentage for OS

    paste(round(median(data$survival_time_os, na.rm = TRUE), 1)),  # Median OS
    round(IQR(data$survival_time_os, na.rm = TRUE), 1)              # IQR OS
  )
  
    # Return the combined results as a dataframe
  return(data.frame(
    Characteristic = Characteristic,
    All_DATA = All_DATA
  ))
  
}


# Generate data for before exclusion
before_df <- merge(
  generate_all_data(combined_clin_before), 
  generate_all_data(combined_clin_sex_before), 
  by = "Characteristic", 
  all = TRUE
)

colnames(before_df)[2] <- "All_Data"
colnames(before_df)[3] <- "Sex_Data"

# Reorder rows to match the original order of 'Characteristic'
order_characteristics <- generate_all_data(combined_clin_before)$Characteristic
before_df <- before_df[match(order_characteristics, before_df$Characteristic), ]

# Generate data for after exclusion
after_df <- merge(
  generate_all_data(filtered_combined_clin), 
  generate_all_data(filtered_combined_clin_sex), 
  by = "Characteristic", 
  all = TRUE
)

colnames(after_df)[2] <- "All_Data"
colnames(after_df)[3] <- "Sex_Data"

# Reorder rows to match the original order of 'Characteristic'
order_characteristics2 <- generate_all_data(combined_clin)$Characteristic
after_df <- after_df[match(order_characteristics2, after_df$Characteristic), ]

rownames(before_df) <-NULL
rownames(after_df) <-NULL

# write.csv(before_df, file = file.path(Base_dir, "outputs/Dataprocessing_outputs", "Summary_of_Data_Before_Exclusion.csv"), row.names = FALSE)
# write.csv(after_df, file = file.path(Base_dir, "outputs/Dataprocessing_outputs", "Summary_of_Data_After_Exclusion.csv"), row.names = FALSE)

# Clean invalid UTF-8 characters in one line for all columns
before_df[] <- lapply(before_df, function(x) if (is.character(x)) iconv(x, to = "UTF-8", sub = " ") else x)
after_df[] <- lapply(after_df, function(x) if (is.character(x)) iconv(x, to = "UTF-8", sub = " ") else x)

# Display the tables using kable
kable(before_df, caption = "Summary of All Data and Data with Sex Before Exclusion", 
      col.names = c("Characteristic", "All Data ", "Data with Sex ")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")

kable(after_df , caption = "Summary of All Data and Data with Sex After Exclusion", 
      col.names = c("Characteristic", "All Data", "Data with Sex")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")

combined_clin <- filtered_combined_clin

```
### After Dataset Exclusion

27 datasets remain after exclusion.

```{r cohorts_display}
# Display all unique cohorts used in the analysis
kable(unique(combined_clin$study), caption = "All cohorts included in the analysis", col.names = "Cohorts") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")
```

## Visulizations for IO datastes 

```{r Visulizations part 2 for IO datasets, width = 12, height = 8}

# 1. Stacked Bar Plot: Treatments by Cancer Type
# Purpose: Visualize the distribution of treatments across cancer types
datasets <- gsub("^ICB_", "", unique(combined_clin$study))
common_datasets <- intersect(summary_table$Dataset, datasets)
summary_table_after <- summary_table[summary_table$Dataset %in% common_datasets, ]
# Find datasets in common between summary_table$Dataset and datasets
common_datasets <- intersect(summary_table$Dataset, datasets)

# Filter summary_table to keep only rows where Dataset is in common_datasets
summary_table_after <- summary_table[summary_table$Dataset %in% common_datasets, ]


# Function to generate a stacked bar plot
generate_treatment_plot <- function(data, title_prefix) {
  treatment_summary <- data %>%
    group_by(Cancer_Type, Treatments) %>%
    summarise(Count = n(), .groups = "drop") %>%
    mutate(Total_Count = sum(Count, na.rm = TRUE), .by = Cancer_Type) %>%
    arrange(desc(Total_Count)) %>%
    mutate(Cancer_Type = factor(Cancer_Type, levels = unique(Cancer_Type)))

  plot <- ggplot(treatment_summary, aes(x = Cancer_Type, y = Count, fill = Treatments)) +
    geom_bar(stat = "identity", position = "stack") +
    geom_text(aes(label = Count), position = position_stack(vjust = 0.5), size = 3) +
    theme_minimal() +
    labs(
      title = paste0(title_prefix, " (", length(unique(data$Dataset)), " datasets)"),
      x = "Cancer Type", y = "Count", fill = "Treatment"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  return(plot)
}

# Generate plots
treatment_bar_plot_full <- generate_treatment_plot(summary_table, "Treatments by Cancer Type (Full Dataset)")
treatment_bar_plot_filtered <- generate_treatment_plot(summary_table_after, "Treatments by Cancer Type (Filtered Dataset)")

# Display plots and captions
treatment_bar_plot_full
treatment_bar_plot_filtered
```



```{r Visulizations part 3 for IO datasets, width = 12, height = 8}

# 2. Boxplot: Age Distribution by Cancer Type
# Purpose: Display the age distribution across cancer types with median and patient counts
create_age_distribution_plot <- function(data, title_suffix) {
  # Filter data for non-numeric ages
  filtered_data <- data %>%
    filter(!is.na(as.numeric(age)))
  
  # Calculate patient counts
  cancer_type_patient_counts <- filtered_data %>%
    group_by(cancer_type) %>%
    summarise(Patient_Count = n(), .groups = "drop")
  
  # Find max age
  max_age <- max(as.numeric(filtered_data$age), na.rm = TRUE)
  
  # Generate plot
  plot <- ggplot(filtered_data, aes(x = cancer_type, y = as.numeric(age))) +
    geom_boxplot(fill = "lightblue") +
    stat_summary(fun = median, geom = "text", aes(label = round(..y.., 1)), vjust = -0.5, color = "darkred", size = 4) +
    geom_text(data = cancer_type_patient_counts, 
              aes(x = cancer_type, y = max_age + 5, label = paste0("n=", Patient_Count)),
              size = 4, fontface = "bold", color = "black") +
    theme_minimal() +
    labs(
      title = paste("Age Distribution by Cancer Type", title_suffix),
      x = "Cancer Type", y = "Age (Years)"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  return(plot)
}

# Generate plots
age_distribution_plot_before <- create_age_distribution_plot(combined_clin_before, "(Before Filtering)")
age_distribution_plot <- create_age_distribution_plot(combined_clin, "(After Filtering)")

# Save plots if needed
# ggsave(filename = file.path(Base_dir, "outputs/Dataprocessing_outputs/Plots/Age_Distribution_by_Cancer_Type_Before_Filtering.png"), 
#        plot = age_distribution_plot_before, bg = "white")
# ggsave(filename = file.path(Base_dir, "outputs/Dataprocessing_outputs/Plots/Age_Distribution_by_Cancer_Type_After_Filtering.png"), 
#        plot = age_distribution_plot, bg = "white")
age_distribution_plot_before
age_distribution_plot

```

```{r Visulizations for IO datasets, width = 15, height = 8}

# 3. Stacked Bar Plot: Patients by Cancer Type and Treatment
# Purpose: Summarize and visualize the number of patients across cancer types and treatments

# Function for summarizing data and generating bar plots
create_summary_plot <- function(data, title_prefix) {
  # Summarize data for the plot
  summary_plot_data <- data %>%
    group_by(Cancer_Type, Treatments) %>%
    summarise(Patients = sum(Patients, na.rm = TRUE), .groups = "drop") %>%
    mutate(Total_Patients = sum(Patients, na.rm = TRUE), .by = Cancer_Type)
  
  # Summarize data and calculate total patients per Cancer Type
  summary_plot_data <- summary_plot_data %>%
    group_by(Cancer_Type) %>%
    mutate(Total_Patients = sum(Patients, na.rm = TRUE)) %>%
    ungroup()
  
  # Reorder Cancer_Type by Total Patients
  summary_plot_data <- summary_plot_data %>%
    mutate(Cancer_Type = factor(Cancer_Type, levels = summary_plot_data %>%
                                  group_by(Cancer_Type) %>%
                                  summarise(Total_Patients = sum(Patients, na.rm = TRUE)) %>%
                                  arrange(desc(Total_Patients)) %>%
                                  pull(Cancer_Type)))
  
  # Create the bar plot ordered by Total Patients
  patient_bar_plot <- ggplot(summary_plot_data, aes(x = Cancer_Type, y = Patients, fill = Treatments)) +
    geom_bar(stat = "identity", position = "stack") +
    theme_minimal() +
    labs(
      title = paste0(title_prefix, " ", length(unique(data$Dataset)), " "),
      x = "Cancer Type", y = "Number of Patients", fill = "Treatment Type"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(size = 14, face = "bold")
    )
  
  return(patient_bar_plot)
}

# Generate plots for both datasets
plot_full <- create_summary_plot(summary_table, "Number of Patients by Cancer Type and Treatment (Full Dataset)")
plot_filtered <- create_summary_plot(summary_table_after, "Number of Patients by Cancer Type and Treatment (Filtered Dataset)")

# Save the plots
# ggsave(filename = file.path(Base_dir, "outputs/Dataprocessing_outputs/Plots/Patients_by_Cancer_Type_and_Treatment_Full.png"), plot = plot_full, bg = "white")
# ggsave(filename = file.path(Base_dir, "outputs/Dataprocessing_outputs/Plots/Patients_by_Cancer_Type_and_Treatment_Filtered.png"), plot = plot_filtered, bg = "white")


plot_full
plot_filtered
```

```{r Pie charts , width = 8.1, height = 5.1}

# Function to create a pie chart
create_pie_chart <- function(data, title, fill_label, fill_column) {
  ggplot(data, aes(x = "", y = Total_Patients, fill = reorder(!!sym(fill_column), -Total_Patients))) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar("y", start = 0) +
    theme_void() +
    geom_text(aes(label = paste0(Percentage, "%\n(n=", Total_Patients, ")")),
              position = position_stack(vjust = 0.5), size = 1.5) +
    labs(title = title, fill = fill_label) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 13),
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 10)
    )
}

# Summarize data for Treatment pie chart (Before Exclusion)
treatment_summary_before <- summary_table %>%
  group_by(Treatments) %>%
  summarise(Total_Patients = sum(Patients), .groups = "drop") %>%
  mutate(Percentage = round((Total_Patients / sum(Total_Patients)) * 100, 2)) %>%
  arrange(desc(Total_Patients))

# Exclude specific treatments
treatment_summary_filtered <- treatment_summary_before %>%
  filter(!Treatments %in% c("chemo", "chemo+targeted", "targeted"))

# Summarize data for Cancer Type pie chart (Before Exclusion)
cancer_type_summary_before <- summary_table %>%
  group_by(Cancer_Type) %>%
  summarise(Total_Patients = sum(Patients), .groups = "drop") %>%
  mutate(Percentage = round((Total_Patients / sum(Total_Patients)) * 100, 2)) %>%
  arrange(desc(Total_Patients))

# Combine smaller cancer types into "Others" for Before Exclusion
total_patients <- sum(cancer_type_summary_before$Total_Patients)
cancer_type_summary_combined <- cancer_type_summary_before %>%
  mutate(
    Cancer_Type = ifelse(
      Cancer_Type %in% c("Esophageal", "Colon", "Lymph_node", "Unknown", "HNC", "Liver"),
      "Others (Esophageal, Colon, Lymph_node, Unknown, HNC, Liver)",
      Cancer_Type
    )
  ) %>%
  group_by(Cancer_Type) %>%
  summarise(Total_Patients = sum(Total_Patients), .groups = "drop") %>%
  mutate(Percentage = round((Total_Patients / total_patients) * 100, 2)) %>%
  arrange(desc(Total_Patients))

# Create Treatment pie chart (Before Exclusion)
treatment_pie_before <- create_pie_chart(treatment_summary_filtered, "Patient Distribution by Treatment (Before Exclusion)", "Treatment Type", "Treatments")

# Create Cancer Type pie chart (Before Exclusion)
cancer_type_pie_before <- create_pie_chart(cancer_type_summary_combined, "Patient Distribution by Cancer Type (Before Exclusion)", "Cancer Type", "Cancer_Type")

# Filter summary_table to keep only rows with datasets in `common_datasets`
excluded_study <- sub("ICB_", "", unique(combined_clin$study))
summary_table_after <- summary_table[summary_table$Dataset %in% excluded_study, ]

# Summarize data for Treatment pie chart (After Exclusion)
treatment_summary_after <- summary_table_after %>%
  group_by(Treatments) %>%
  summarise(Total_Patients = sum(Patients), .groups = "drop") %>%
  mutate(Percentage = round((Total_Patients / sum(Total_Patients)) * 100, 2)) %>%
  arrange(desc(Total_Patients))

# Summarize data for Cancer Type pie chart (After Exclusion)
cancer_type_summary_after <- summary_table_after %>%
  group_by(Cancer_Type) %>%
  summarise(Total_Patients = sum(Patients), .groups = "drop") %>%
  mutate(Percentage = round((Total_Patients / sum(Total_Patients)) * 100, 2)) %>%
  arrange(desc(Total_Patients))

# Combine smaller cancer types into "Others" for After Exclusion
total_patients_after <- sum(cancer_type_summary_after$Total_Patients)
cancer_type_summary_combined_after <- cancer_type_summary_after %>%
  mutate(
    Cancer_Type = ifelse(
      Cancer_Type %in% c("Esophageal", "Pancreas", "Gastric", "Brain", "Ureteral"),
      "Others (Esophageal, Colon, Lymph_node, Unknown, HNC, Liver)",
      Cancer_Type
    )
  ) %>%
  group_by(Cancer_Type) %>%
  summarise(Total_Patients = sum(Total_Patients), .groups = "drop") %>%
  mutate(
    Percentage = round((Total_Patients / total_patients_after) * 100, 2)
  ) %>%
  arrange(desc(Total_Patients))

# Create Treatment pie chart (After Exclusion)
treatment_pie_after <- create_pie_chart(treatment_summary_after, "Patient Distribution by Treatment (After Exclusion)", "Treatment Type", "Treatments")

# Create Cancer Type pie chart (After Exclusion)
cancer_type_pie_after <- create_pie_chart(cancer_type_summary_combined_after, "Patient Distribution by Cancer Type (After Exclusion)", "Cancer Type", "Cancer_Type")

# Display pie charts\
print(treatment_pie_before)
print(treatment_pie_after)
print(cancer_type_pie_before)
print(cancer_type_pie_after)

```


```{r Snakey after for IO datasets}

# Function to generate a Sankey diagram with percentages
generate_sankey_with_percentages <- function(data) {
  # Calculate total patients
  total_patients <- sum(data$Patients, na.rm = TRUE)
  
  # Summarize data for Sankey diagram
  summary_sankey <- data %>%
    group_by(Cancer_Type, Treatments, Clinical_Endpoints) %>%
    summarise(Total_Patients = sum(Patients, na.rm = TRUE), .groups = "drop")
  
  # Prepare Nodes with counts and percentages
  nodes <- data.frame(
    name = c(
      paste0("Total Patients\n(n=", total_patients, ", 100%)"),
      paste0(unique(summary_sankey$Cancer_Type), "\n(n=", 
             sapply(unique(summary_sankey$Cancer_Type), function(ct) sum(summary_sankey$Total_Patients[summary_sankey$Cancer_Type == ct])), 
             ", ", 
             round(sapply(unique(summary_sankey$Cancer_Type), function(ct) 
               sum(summary_sankey$Total_Patients[summary_sankey$Cancer_Type == ct]) / total_patients * 100), 2), 
             "%)"),
      paste0(unique(summary_sankey$Treatments), "\n(n=", 
             sapply(unique(summary_sankey$Treatments), function(t) sum(summary_sankey$Total_Patients[summary_sankey$Treatments == t])), 
             ", ", 
             round(sapply(unique(summary_sankey$Treatments), function(t) 
               sum(summary_sankey$Total_Patients[summary_sankey$Treatments == t]) / total_patients * 100), 2), 
             "%)"),
      paste0(unique(summary_sankey$Clinical_Endpoints), "\n(n=", 
             sapply(unique(summary_sankey$Clinical_Endpoints), function(ce) sum(summary_sankey$Total_Patients[summary_sankey$Clinical_Endpoints == ce])), 
             ", ", 
             round(sapply(unique(summary_sankey$Clinical_Endpoints), function(ce) 
               sum(summary_sankey$Total_Patients[summary_sankey$Clinical_Endpoints == ce]) / total_patients * 100), 2), 
             "%)")
    )
  )
  
  # Prepare Links
  # First layer: Total Patients to Cancer Types
  links <- summary_sankey %>%
    group_by(Cancer_Type) %>%
    summarise(Total_Patients = sum(Total_Patients)) %>%
    mutate(
      source = 0, # First node (Total Patients)
      target = match(paste0(Cancer_Type, "\n(n=", 
                            sapply(Cancer_Type, function(ct) sum(summary_sankey$Total_Patients[summary_sankey$Cancer_Type == ct])), 
                            ", ", 
                            round(sapply(Cancer_Type, function(ct) 
                              sum(summary_sankey$Total_Patients[summary_sankey$Cancer_Type == ct]) / total_patients * 100), 2), 
                            "%)"), nodes$name) - 1,
      value = Total_Patients
    ) %>%
    select(source, target, value)
  
  # Second layer: Cancer Types to Treatments
  links <- links %>%
    bind_rows(
      summary_sankey %>%
        mutate(
          source = match(paste0(Cancer_Type, "\n(n=", 
                                sapply(Cancer_Type, function(ct) sum(summary_sankey$Total_Patients[summary_sankey$Cancer_Type == ct])), 
                                ", ", 
                                round(sapply(Cancer_Type, function(ct) 
                                  sum(summary_sankey$Total_Patients[summary_sankey$Cancer_Type == ct]) / total_patients * 100), 2), 
                                "%)"), nodes$name) - 1,
          target = match(paste0(Treatments, "\n(n=", 
                                sapply(Treatments, function(t) sum(summary_sankey$Total_Patients[summary_sankey$Treatments == t])), 
                                ", ", 
                                round(sapply(Treatments, function(t) 
                                  sum(summary_sankey$Total_Patients[summary_sankey$Treatments == t]) / total_patients * 100), 2), 
                                "%)"), nodes$name) - 1,
          value = Total_Patients
        ) %>%
        select(source, target, value)
    )
  
  # Third layer: Treatments to Clinical Endpoints
  links <- links %>%
    bind_rows(
      summary_sankey %>%
        mutate(
          source = match(paste0(Treatments, "\n(n=", 
                                sapply(Treatments, function(t) sum(summary_sankey$Total_Patients[summary_sankey$Treatments == t])), 
                                ", ", 
                                round(sapply(Treatments, function(t) 
                                  sum(summary_sankey$Total_Patients[summary_sankey$Treatments == t]) / total_patients * 100), 2), 
                                "%)"), nodes$name) - 1,
          target = match(paste0(Clinical_Endpoints, "\n(n=", 
                                sapply(Clinical_Endpoints, function(ce) sum(summary_sankey$Total_Patients[summary_sankey$Clinical_Endpoints == ce])), 
                                ", ", 
                                round(sapply(Clinical_Endpoints, function(ce) 
                                  sum(summary_sankey$Total_Patients[summary_sankey$Clinical_Endpoints == ce]) / total_patients * 100), 2), 
                                "%)"), nodes$name) - 1,
          value = Total_Patients
        ) %>%
        select(source, target, value)
    )
  
  # Create Sankey Diagram
  sankey <- sankeyNetwork(
    Links = links, Nodes = nodes,
    Source = "source", Target = "target",
    Value = "value", NodeID = "name",
    units = "Patients", fontSize = 12, nodeWidth = 30,
    nodePadding = 10, width = 1050, height = 600
  )
  
  # Return the Sankey Diagram for visualization in the session
  return(sankey)
}

# Usage Example
sankey_before <- generate_sankey_with_percentages(summary_table)
sankey_after <- generate_sankey_with_percentages(summary_table_after)

sankey_before
sankey_after

#saveNetwork(sankey_before, file = "~/BHK lab/EcoTyper_Project/outputs/Dataprocessing_outputs/Plots/Sankey_Cancer_Treatments_Endpoints_Before.html")
# saveNetwork(sankey_after, file = "~/BHK lab/EcoTyper_Project/outputs/Dataprocessing_outputs/Plots/Sankey_Cancer_Treatments_Endpoints_After.html")

```

## Detailed clinical characteristics of the IO datasets

processes clinical data from multiple datasets, extracts key characteristics, and formats them into a summary table.

```{r Table 2}

# Initialize the results list
results <- list()

# Loop through unique studies in combined_clin
for (study_name in unique(combined_clin$study)) {
  
  # Subset the data for the current study
  study_data <- combined_clin[combined_clin$study == study_name, ]
  
  # Number of samples
  n_samples <- nrow(study_data)
  
  # Age calculations
  study_data$age <- as.numeric(as.character(study_data$age))
  valid_ages <- na.omit(study_data$age)
  age_median <- if(length(valid_ages) > 0) round(median(valid_ages, na.rm = TRUE), 1) else NA
  age_iqr <- if(length(valid_ages) > 1) round(IQR(valid_ages, na.rm = TRUE), 1) else NA
  age_na <- sum(is.na(study_data$age))
  
  # Sex calculations
  n_female <- sum(study_data$sex == "F", na.rm = TRUE)
  n_male <- sum(study_data$sex == "M", na.rm = TRUE)
  sex_na <- sum(is.na(study_data$sex))
  
  # Cancer type (assuming the cancer type is consistent within the study)
  cancer_type <- unique(study_data$cancer_type)
  
  # Treatment type
  treatment_type <- paste(unique(study_data$treatment), collapse = ", ")
  
  # Response counts
  no_response <- sum(study_data$response == "NR", na.rm = TRUE)
  response <- sum(study_data$response == "R", na.rm = TRUE)
  
  # PFS counts
  no_progression <- sum(study_data$event_occurred_pfs == 0, na.rm = TRUE)
  progression <- sum(study_data$event_occurred_pfs == 1, na.rm = TRUE)
  pfs_median <- if (!all(is.na(study_data$survival_time_pfs))) round(median(na.omit(study_data$survival_time_pfs), na.rm = TRUE), 1) else NA
  pfs_iqr <- if (!all(is.na(study_data$survival_time_pfs))) round(IQR(na.omit(study_data$survival_time_pfs), na.rm = TRUE), 1) else NA
  
  # OS counts
  alive <- sum(study_data$event_occurred_os == 0, na.rm = TRUE)
  dead <- sum(study_data$event_occurred_os == 1, na.rm = TRUE)
  os_median <- if (!all(is.na(study_data$survival_time_os))) round(median(na.omit(study_data$survival_time_os), na.rm = TRUE), 1) else NA
  os_iqr <- if (!all(is.na(study_data$survival_time_os))) round(IQR(na.omit(study_data$survival_time_os), na.rm = TRUE), 1) else NA
  
  expression_type <- ifelse(any(study_data$rna == "rnaseq"), "TPM", ifelse(any(study_data$rna == "microarray"), "log2gNor", TPM))
  
  # Compile the results for this study
  results[[study_name]] <- data.frame(
    Study = study_name, 
    N = n_samples, 
    Expression = expression_type, 
    Age_Median = age_median, 
    Age_IQR = age_iqr, 
    Age_NA = age_na, 
    Female = n_female, 
    Male = n_male, 
    Sex_NA = sex_na, 
    Cancer_Type = cancer_type, 
    Treatment_Type = treatment_type, 
    No_Response = no_response, 
    Response = response, 
    No_Progression = no_progression, 
    Progressed = progression, 
    PFS_Median = pfs_median, 
    PFS_IQR = pfs_iqr, 
    Alive = alive, 
    Dead = dead, 
    OS_Median = os_median, 
    OS_IQR = os_iqr,
    stringsAsFactors = FALSE
  )
}

# Combine all results into a single dataframe
results_df <- do.call(rbind, results)

kable(results_df, caption = "Detailed clinical characteristics of the IO datasets") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")

#write.csv(results_df, file = file.path(Base_dir, "outputs/Dataprocessing_outputs", "Detailed_clinical_characteristics_of_the_IO_datasets.csv"), row.names = FALSE)

```

## Table S4: Detailed clinical characteristics of the IO dataset stratify by sex 																									
Processes clinical data stratified by sex from multiple datasets, extracts key characteristics, and formats them into a summary table.

```{r Table 3}

# Initialize the results list for combined_clin_sex
results_sex <- list()

# Loop through unique studies in combined_clin_sex
for (study_name in unique(combined_clin_sex$study)) {
  
  # Subset the data for the current study
  study_data <- combined_clin_sex[combined_clin_sex$study == study_name, ]
  
  # Number of samples
  n_samples <- nrow(study_data)
  
  # Age calculations
  study_data$age <- as.numeric(as.character(study_data$age))
  valid_ages <- na.omit(study_data$age)
  age_median <- if(length(valid_ages) > 0) round(median(valid_ages, na.rm = TRUE), 1) else NA
  age_iqr <- if(length(valid_ages) > 1) round(IQR(valid_ages, na.rm = TRUE), 1) else NA
  age_na <- sum(is.na(study_data$age))
  
  # Sex calculations
  n_female <- sum(study_data$sex == "F", na.rm = TRUE)
  n_male <- sum(study_data$sex == "M", na.rm = TRUE)
  sex_na <- sum(is.na(study_data$sex))
  
  # Cancer type (assuming the cancer type is consistent within the study)
  cancer_type <- unique(study_data$cancer_type)
  
  # Treatment type
  treatment_type <- paste(unique(study_data$treatment), collapse = ", ")
  
  # Response counts
  no_response <- sum(study_data$response == "NR", na.rm = TRUE)
  response <- sum(study_data$response == "R", na.rm = TRUE)
  
  # PFS counts
  no_progression <- sum(study_data$event_occurred_pfs == 0, na.rm = TRUE)
  progression <- sum(study_data$event_occurred_pfs == 1, na.rm = TRUE)
  pfs_median <- if (!all(is.na(study_data$survival_time_pfs))) round(median(na.omit(study_data$survival_time_pfs), na.rm = TRUE), 1) else NA
  pfs_iqr <- if (!all(is.na(study_data$survival_time_pfs))) round(IQR(na.omit(study_data$survival_time_pfs), na.rm = TRUE), 1) else NA
  
  # OS counts
  alive <- sum(study_data$event_occurred_os == 0, na.rm = TRUE)
  dead <- sum(study_data$event_occurred_os == 1, na.rm = TRUE)
  os_median <- if (!all(is.na(study_data$survival_time_os))) round(median(na.omit(study_data$survival_time_os), na.rm = TRUE), 1) else NA
  os_iqr <- if (!all(is.na(study_data$survival_time_os))) round(IQR(na.omit(study_data$survival_time_os), na.rm = TRUE), 1) else NA
  
  # Compile the results for this study
  results_sex[[study_name]] <- data.frame(
    Study = study_name, 
    N = n_samples, 
    Expression = expression_type, 
    Age_Median = age_median, 
    Age_IQR = age_iqr, 
    Age_NA = age_na, 
    Female = n_female, 
    Male = n_male, 
    Sex_NA = sex_na, 
    Cancer_Type = cancer_type, 
    Treatment_Type = treatment_type, 
    No_Response = no_response, 
    Response = response, 
    No_Progression = no_progression, 
    Progressed = progression, 
    PFS_Median = pfs_median, 
    PFS_IQR = pfs_iqr, 
    Alive = alive, 
    Dead = dead, 
    OS_Median = os_median, 
    OS_IQR = os_iqr,
    stringsAsFactors = FALSE
  )}
  
# Combine all results into a single dataframe
results_df_sex <- do.call(rbind, results_sex)

# Display 
kable(results_df_sex, caption = "Detailed clinical characteristics of the IO datasets") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")

# write.csv(results_df_sex, file = file.path(Base_dir, "outputs/Dataprocessing_outputs", "Detailed_clinical_characteristics_of_the_IO_datasets_with_sex.csv"), row.names = FALSE)

```


